%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Extensions of Schur Numbers
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
¥documentclass[11pt]{article}

¥usepackage[margin=1in]{geometry}
¥usepackage{amsmath,amssymb,amsthm}
¥usepackage{mathtools}
¥usepackage{hyperref}
¥usepackage{booktabs}
¥usepackage{microtype}
¥usepackage{enumitem}

¥theoremstyle{plain}
¥newtheorem{theorem}{Theorem}[section]
¥newtheorem{lemma}[theorem]{Lemma}
¥newtheorem{proposition}[theorem]{Proposition}
¥newtheorem{corollary}[theorem]{Corollary}
¥newtheorem{conjecture}[theorem]{Conjecture}

¥theoremstyle{definition}
¥newtheorem{definition}[theorem]{Definition}
¥newtheorem{remark}[theorem]{Remark}
¥newtheorem{observation}[theorem]{Observation}

¥newcommand{¥ZZ}{¥mathbb{Z}}
¥newcommand{¥DS}{¥mathrm{DS}}
¥newcommand{¥MS}{¥mathrm{MS}}

¥begin{document}

¥title{Extensions of Schur Numbers:¥¥
Order-Invariance, Density Thresholds, and Multiplicative Variants}

¥author{Alex Towell¥thanks{Department of Computer Science,
Southern Illinois University Edwardsville.
Email: ¥texttt{atowell@siue.edu}}}

¥date{}
¥maketitle

¥begin{abstract}
We report computational results in three directions extending
classical Schur numbers.
First, we study Schur numbers $S(G,k)$ for finite abelian groups‾$G$
and find that $S(G,1)$ and $S(G,2)$ depend only on $|G|$, verified
through order‾20, while $S(G,3)$ is ¥emph{not} order-invariant:
$S(¥ZZ/9¥ZZ,3)=8$ but $S(¥ZZ/3¥ZZ¥times¥ZZ/3¥ZZ,3)=7$.
Second, we introduce density Schur numbers $¥DS(k,¥alpha)$ and
determine $¥DS(2,¥tfrac{1}{2})=5$ with a Lean‾4 proof,
identifying a phase transition in the parameter‾$¥alpha$.
Third, we tabulate Rado numbers for several linear equations and
compute multiplicative Schur numbers $¥MS(k)$ on the domain
$¥{2,¥ldots,N¥}$, finding $¥MS(1)=3$ and $¥MS(2)=31$.
All results are computational (exhaustive enumeration or SAT-based)
and should be regarded as such until independent formal verification.
¥end{abstract}

%% ================================================================
¥section{Schur numbers in finite abelian groups}¥label{sec:groups}
%% ================================================================

¥subsection{Definitions}

Let $G$ be a finite abelian group written additively.
A subset $C¥subseteq G$ is ¥emph{sum-free} if there are no
$a,b,c¥in C$ with $a+b=c$ (the case $a=b$ is permitted, so
$¥{0¥}$ is never sum-free since $0+0=0$).

¥begin{definition}
$S(G,k)$ is the maximum size of a subset $A¥subseteq G$ admitting a
partition $A=C_1¥sqcup¥cdots¥sqcup C_k$ into $k$ sum-free sets.
¥end{definition}

¥noindent
Equivalently, $S(G,k)+1$ is the minimum size of a subset of‾$G$ that
forces a monochromatic Schur triple in every $k$-coloring.
When $G=¥ZZ/n¥ZZ$ and $k=1$, the quantity $S(G,1)$ is the maximum
sum-free subset of‾$¥ZZ/n¥ZZ$.

For a single color ($k=1$), Green and
Ruzsa‾¥cite{green-ruzsa} established asymptotically tight bounds on
the maximum sum-free subset of a general abelian group.
For $¥ZZ/p¥ZZ$ with $p$ an odd prime, the interval
$¥{x : p/3 < x < 2p/3¥}$ is a maximum sum-free set of size
$¥lfloor p/3 ¥rfloor + O(1)$; we have verified this computationally
for all primes $p ¥le 19$.

¥subsection{Order-invariance}

We say that $S(-,k)$ is ¥emph{order-invariant} if $S(G,k)$
depends only on $|G|$ for all finite abelian groups‾$G$.
Since any group isomorphism preserves sum-freeness, $S(G,k)$
is certainly an isomorphism invariant.
The question is whether non-isomorphic groups of the same order can
differ.

¥begin{observation}[Computational, verified through order 20]¥label{obs:inv}
$S(G,1)$ and $S(G,2)$ are order-invariant for finite abelian groups
of order $¥le 20$.
This covers 30 non-isomorphic groups, including all‾5 abelian
groups of order‾16 (which all satisfy $S(G,1)=8$ and $S(G,2)=12$).
¥end{observation}

¥noindent
For $k=1$, order-invariance may follow from the Green--Ruzsa
theory‾¥cite{green-ruzsa}, which determines $S(G,1)$ in terms of
$|G|$ and the group's structure constants.
The $k=2$ invariance appears to be a new observation.

¥begin{observation}[Computational]¥label{obs:break}
$S(G,3)$ is ¥emph{not} order-invariant.
The first failure occurs at order‾$9$:
¥[
S(¥ZZ/9¥ZZ,¥, 3) = 8, ¥qquad
S(¥ZZ/3¥ZZ¥times¥ZZ/3¥ZZ,¥, 3) = 7.
¥]
A second failure occurs at order‾12:
$S(¥ZZ/3¥ZZ¥times¥ZZ/4¥ZZ,¥, 3) = 11$ while
$S(¥ZZ/2¥ZZ¥times¥ZZ/2¥ZZ¥times¥ZZ/3¥ZZ,¥, 3) = 10$.
¥end{observation}

¥noindent
The underlying mechanism appears related to the group exponent:
$¥ZZ/9¥ZZ$ has exponent‾$9$ while
$(¥ZZ/3¥ZZ)^2$ has exponent‾$3$, and the higher exponent allows more
room for three independent sum-free color classes.

¥subsection{Tables}

Table‾¥ref{tab:cyclic} records $S(¥ZZ/n¥ZZ, k)$ for small‾$n$ and
$k=1,2,3$.

¥begin{table}[ht]
¥centering
¥caption{Schur numbers $S(¥ZZ/n¥ZZ,k)$ for $n=2,¥ldots,16$.
All values computed by exhaustive enumeration.}¥label{tab:cyclic}
¥smallskip
¥begin{tabular}{@{}r rrr r@{}}
¥toprule
$n$ & $S(-,1)$ & $S(-,2)$ & $S(-,3)$ & $S_2/n$ ¥¥
¥midrule
 2  &  1 &  1 &  1 & 0.500 ¥¥
 3  &  1 &  2 &  2 & 0.667 ¥¥
 4  &  2 &  3 &  3 & 0.750 ¥¥
 5  &  2 &  4 &  4 & 0.800 ¥¥
 6  &  3 &  4 &  5 & 0.667 ¥¥
 7  &  2 &  4 &  6 & 0.571 ¥¥
 8  &  4 &  6 &  7 & 0.750 ¥¥
 9  &  3 &  6 &  8 & 0.667 ¥¥
10  &  5 &  8 &  9 & 0.800 ¥¥
11  &  4 &  8 & 10 & 0.727 ¥¥
12  &  6 &  9 &    & 0.750 ¥¥
13  &  4 &  8 &    & 0.615 ¥¥
14  &  7 &  9 &    & 0.643 ¥¥
15  &  6 & 12 &    & 0.800 ¥¥
16  &  8 & 12 &    & 0.750 ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥noindent
Blank entries in the $k=3$ column indicate that the exhaustive
computation ($3^n$ colorings) was not run for those‾$n$ due to time
constraints.
For $k=3$, note that $S(¥ZZ/n¥ZZ,3) = n-1$ for all $n ¥le 11$:
the $n-1$ nonzero elements of $¥ZZ/n¥ZZ$ can always be 3-colored
with each color class sum-free, at least through $n=11$.

¥begin{conjecture}
$S(G,2)$ is order-invariant for all finite abelian groups.
¥end{conjecture}

%% ================================================================
¥section{Density Schur numbers}¥label{sec:density}
%% ================================================================

¥begin{definition}¥label{def:ds}
For integers $k¥ge 1$ and a real parameter $¥alpha¥in[0,1]$,
define the ¥emph{density Schur number} $¥DS(k,¥alpha)$ as the
minimum‾$N$ such that every $k$-coloring of $¥{1,¥ldots,N¥}$ has at
least one color class‾$C_i$ that is ¥emph{good}: either $C_i$
contains a Schur triple ($a+b=c$ with $a,b,c¥in C_i$ and
$a,b,c¥ge 1$) or $|C_i| > ¥alpha N$.
¥end{definition}

¥begin{remark}[Indexing caveat]¥label{rem:index}
Definition‾¥ref{def:ds} uses the 1-indexed domain $¥{1,¥ldots,N¥}$,
which is the natural setting for integer Schur triples.
An alternative 0-indexed definition on $¥{0,¥ldots,N-1¥}$ would give
different values because $0+0=0$ is a ``free'' Schur triple that
forces any color class containing‾0 to be good by default.
Our Lean‾4 formalization‾¥cite{code-repo} uses the
1-indexed convention (elements $¥{1,¥ldots,N¥}$, represented
internally as ¥texttt{Fin‾N} with the Schur condition
$(a+1)+(b+1)=(c+1)$).
¥end{remark}

¥begin{theorem}[Lean-verified]¥label{thm:ds2}
$¥DS(2,¥tfrac{1}{2}) = 5$.
That is, for every 2-coloring of $¥{1,2,3,4,5¥}$, some
color class either contains a Schur triple or has more than
$¥tfrac{1}{2}¥cdot 5 = 2.5$ elements.
The witness coloring on $¥{1,2,3,4¥}$ that avoids this is
$C_0=¥{1,4¥}$, $C_1=¥{2,3¥}$: both classes are sum-free and both
have exactly $2 ¥le ¥lfloor 4/2 ¥rfloor$ elements, so neither
exceeds the density threshold.
¥end{theorem}

¥noindent
The proof is formalized in Lean‾4 with zero sorry statements.
The lower bound (exhibiting the witness coloring) and upper bound
(pigeonhole: some class has $¥ge 3 > 5/2$ elements) are both
machine-checked.

¥begin{theorem}[Pigeonhole]¥label{thm:ds-pigeonhole}
$¥DS(k,¥, 1/k) = S(k)+1$,
where $S(k)$ is the classical Schur number.
¥end{theorem}

¥begin{proof}
At $N = S(k)+1$, every $k$-coloring of $¥{1,¥ldots,N¥}$ forces a
monochromatic Schur triple by the definition of‾$S(k)$, so every
color class is good (the Schur-triple condition alone suffices).
At $N = S(k)$, there exists a $k$-coloring with no monochromatic
Schur triple.
By pigeonhole, some class has $¥ge ¥lceil N/k ¥rceil$ elements, but
the density condition requires ¥emph{strictly} more than $N/k$
elements; when $k$ divides‾$N$ and the partition is balanced, this
threshold is not exceeded.
¥end{proof}

¥subsection{Phase transition for $k=2$}

A related quantity is the ¥emph{density Ramsey number}
$¥DS'(k,¥alpha)$: the minimum‾$N$ such that every $k$-coloring of
$¥{1,¥ldots,N¥}$ has a color class‾$C$ satisfying ¥emph{both}
$|C|¥ge¥alpha N$ ¥emph{and} $C$ contains a Schur triple (the
conjunction, not the disjunction in Definition‾¥ref{def:ds}).
Computational experiments reveal three regimes for
$¥DS'(2,¥alpha)$:

¥begin{enumerate}[nosep]
¥item ¥textbf{Regime‾1} ($¥alpha ¥le 0.59$): $¥DS'(2,¥alpha) = 5$,
matching $S(2)+1$.
¥item ¥textbf{Regime‾2} ($0.60 ¥le ¥alpha ¥le 0.66$):
$¥DS'(2,¥alpha) = 6$.
¥item ¥textbf{Regime‾3} ($¥alpha ¥ge 0.67$): $¥DS'(2,¥alpha) > 12$.
¥end{enumerate}

¥noindent
The transition from Regime‾1 to Regime‾2 is sharp: at $¥alpha=0.60$,
$N=5$ has an avoiding coloring (one class can have 3 out of 5
elements, and $3 < 0.60 ¥times 5 = 3.0$, so the density condition
is not triggered), but $N=6$ does not.
These values are determined by exhaustive enumeration over all
$2^N$ colorings.

%% ================================================================
¥section{Rado numbers and multiplicative Schur}¥label{sec:rado}
%% ================================================================

¥subsection{Rado numbers for linear equations}

A Rado number $R(E,k)$ for an equation‾$E$ is the minimum‾$N$ such
that every $k$-coloring of $¥{1,¥ldots,N¥}$ has a monochromatic
solution to‾$E$.
We computed the following values using a SAT-based approach
(CaDiCaL 1.5 solver with direct encoding).

¥begin{table}[ht]
¥centering
¥caption{Rado numbers $R(E,k)$ for small equations and
$k=1,2,3$.}¥label{tab:rado}
¥smallskip
¥begin{tabular}{@{}l rrr@{}}
¥toprule
Equation $E$ & $k=1$ & $k=2$ & $k=3$ ¥¥
¥midrule
$a+b=c$ (Schur)       &  2 &  5 & 14 ¥¥
$a+b=2c$ (van der Waerden) &  --- &  9 & 27 ¥¥
$a+b+c=d$             &  --- & 11 & 43 ¥¥
$a+2b=3c$             &  --- & 13 & --- ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥noindent
The Schur values $R(a{+}b{=}c,¥, k) = S(k)+1$ are well-known
($S(1)=1$, $S(2)=4$, $S(3)=13$, $S(4)=44$).
The van der Waerden values match $W(k;3)$.
The values $R(a{+}b{+}c{=}d,¥, 2)=11$ and
$R(a{+}b{+}c{=}d,¥, 3)=43$ are verified by our SAT solver and are
consistent with the known literature on Rado's theorem.
A dash (---) indicates that the value was not computed (either
trivial or exceeding our search bound).

¥subsection{Multiplicative Schur numbers}

¥begin{definition}
$¥MS(k)$ is the minimum‾$N$ such that every $k$-coloring of
$¥{2,3,¥ldots,N¥}$ contains a monochromatic ¥emph{multiplicative
triple} $¥{a,b,ab¥}$ (with $a¥le b$).
¥end{definition}

¥noindent
The domain starts at‾2 to exclude‾1, which would trivially satisfy
$1¥cdot a = a$ for all‾$a$.

¥begin{observation}[SAT-verified]¥label{obs:ms}
$¥MS(1) = 3$ and $¥MS(2) = 31$.
Equivalently, $¥{2,3¥}$ can be 1-colored avoiding multiplicative
triples, but $¥{2,3,4¥}$ cannot (since $2¥cdot 2=4$).
For $k=2$, $¥{2,¥ldots,31¥}$ can be 2-colored, but
$¥{2,¥ldots,32¥}$ cannot.
¥end{observation}

¥noindent
Additionally, SAT verification for $k=3$ yields
$¥MS(3)=16383=2^{14}-1$ (with $¥{2,¥ldots,16383¥}$ admitting a valid
3-coloring but $¥{2,¥ldots,16384¥}$ not).
This suggests the formula
¥begin{equation}¥label{eq:ms-formula}
¥MS(k) = 2^{(3^k+1)/2} - 1,
¥end{equation}
which gives $¥MS(1)=2^2-1=3$, $¥MS(2)=2^5-1=31$, $¥MS(3)=2^{14}-1=16383$.
We emphasize that ¥eqref{eq:ms-formula} is a ¥emph{conjecture}
extrapolated from three data points, not a theorem.
The exponents $2,5,14$ satisfy the recurrence $f(k)=3f(k-1)-1$ with
$f(1)=2$.

Multiplicative Schur triples have been studied recently by
Mattos, Mergoni Cecchelli, and Parczyk‾¥cite{mattos2024}, who
investigate the equation $xy=z$ in the integers and obtain asymptotic
density bounds.
Our exact small-$k$ values may complement their asymptotic results,
but we have not verified whether the values $¥MS(1)$ and $¥MS(2)$
appear in prior work.

¥subsection{Mixed additive-multiplicative constraints}

One can also forbid ¥emph{both} additive Schur triples
$¥{a,b,a{+}b¥}$ and multiplicative triples $¥{a,b,ab¥}$
simultaneously.
We find that the resulting mixed number equals $S(k)+1$ for
$k=1,2,3$: the additive constraint completely dominates.
This is consistent with the much larger values of $¥MS(k)$ compared
to $S(k)+1$ (e.g., $¥MS(2)=31$ vs.¥ $S(2)+1=5$).

%% ================================================================
¥section*{Acknowledgments}
%% ================================================================

Computational analysis performed with AI assistance (Claude,
Anthropic).
All SAT-based results verified independently using CaDiCaL.
The Lean‾4 formalization of $¥DS(2,¥tfrac{1}{2})=5$ was
machine-checked with zero sorry statements.
Source code and Lean proofs are available at‾¥cite{code-repo}.

%% ================================================================
¥begin{thebibliography}{9}
%% ================================================================

¥bibitem{green-ruzsa}
B.‾Green and I.‾Ruzsa,
``Sum-free sets in abelian groups,''
¥emph{Israel J.¥ Math.} ¥textbf{147} (2005), 157--188.

¥bibitem{mattos2024}
L.‾Mattos, T.‾Mergoni Cecchelli, and J.‾Parczyk,
``On product Schur triples in the integers,''
¥emph{SIAM J.¥ Discrete Math.} (2024/2025).
¥url{https://epubs.siam.org/doi/10.1137/24M1632875}

¥bibitem{schur-survey}
W.‾Deuber, ``Developments based on Rado's dissertation
¥emph{Studien zur Kombinatorik},''
in ¥emph{Surveys in Combinatorics} (1995), 52--74.

¥bibitem{code-repo}
A.‾Towell,
``Erd¥H{o}s problems: computational experiments,''
2026.
Zenodo code deposit.
DOI: ¥texttt{10.5281/zenodo.18638635}.

¥end{thebibliography}

¥end{document}
